---
title: "Genomic Predictors of Duplication Mode in *Hydra vulgaris* AEP"
subtitle: "Exploratory Data Analysis"
author: "Rachel"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme: darkly
    highlight-style: dracula
    fig-width: 9
    fig-height: 5
    embed-resources: true
    number-sections: true
execute:
  warning: false
  message: false
  cache: true
---

## Background and Motivating Question

Gene duplication is a primary mechanism of genome evolution, and the *mode* of 
duplication — tandem, proximal, segmental, or dispersed — has profound consequences 
for how duplicated gene copies diverge in function. In cnidarians like *Hydra*, 
opsin genes (light-sensitive G protein-coupled receptors mediating phototransduction) 
have undergone substantial expansion, but it is not known whether this expansion 
occurred preferentially through tandem duplication events or through dispersed 
mechanisms.

**Motivating question:** Do genomic features — gene length, local gene density, 
distance to the nearest gene, and scaffold size — predict whether a gene duplicate 
pair arose via tandem versus dispersed duplication? And are opsin genes 
preferentially involved in tandem duplications compared to the genomic background?

This question matters because:

1. **Tandem duplicates** tend to retain similar regulatory environments and are 
   more likely to undergo concerted evolution or subfunctionalization within 
   a tight genomic cluster.
2. **Opsin gene clusters** in other animals (e.g., vertebrate visual opsins) 
   are often organized as tandem arrays, suggesting tandem duplication may be 
   a conserved feature of opsin family expansion.
3. Understanding duplication mode in *Hydra* opsins informs hypotheses about 
   how photoreceptor diversity arose in early-diverging animals lacking 
   a centralized nervous system.

---

## Setup

```{r setup}
library(tidyverse)
library(pROC)
library(ggeffects)
library(patchwork)
library(broom)
library(rstanarm)
library(bayesplot)
library(GenomicRanges)
library(rtracklayer)
library(Biostrings)
library(scales)
library(ggridges)
library(knitr)
library(kableExtra)
library(viridis)

# ── Shared plot theme ─────────────────────────────────────────────────────────
theme_hydra <- theme_minimal(base_size = 11) +
  theme(
    plot.background   = element_rect(fill = "#0d1117", color = NA),
    panel.background  = element_rect(fill = "#1c2333", color = NA),
    panel.grid        = element_line(color = "#21262d", linewidth = 0.3),
    text              = element_text(color = "#c9d1d9"),
    axis.text         = element_text(color = "#8b949e"),
    axis.title        = element_text(color = "#c9d1d9"),
    plot.title        = element_text(color = "#e6edf3", face = "bold", size = 12),
    plot.subtitle     = element_text(color = "#6e7681", size = 9),
    legend.background = element_rect(fill = "#1c2333", color = NA),
    legend.text       = element_text(color = "#c9d1d9"),
    plot.caption      = element_text(color = "#6e7681", size = 8),
    strip.text        = element_text(color = "#c9d1d9", face = "bold")
  )

COLORS <- c(TD = "#ff6b6b", DD = "#4ecdc4", PD = "#ffe66d", SD = "#c77dff")
```

---

## Data Sources

| File | Description | Source |
|---|---|---|
| `dup_pairsAEP.tsv` | 50,364 duplicate gene pairs with type annotation | MCScanX / *Hydra* AEP genome |
| `HVAEP.GeneModels.gff3.gz` | Gene model coordinates | NHGRI HydraAEP portal |
| `HVAEP.genome.fa` | Chromosome-level genome assembly | NHGRI HydraAEP portal |
| `AEP.txt` | List of 50 opsin gene IDs | Lab annotation |

```{r load-data}
# ── Raw duplicate pairs ───────────────────────────────────────────────────────
dup_pairs <- read_tsv("dup_pairsAEP.tsv", col_types = cols()) %>%
  rename(dup1 = 1, dup2 = 2, dup_type = 3) %>%
  filter(!is.na(dup_type), dup_type != "type") %>%
  mutate(dup_type = factor(dup_type, levels = c("TD","PD","SD","DD")))

# ── Gene-level feature table (produced by AEP_gene_models.R) ─────────────────
gene_features <- read_tsv("gene_feature_table_AEP.tsv", col_types = cols())

# ── Pair-level model table ────────────────────────────────────────────────────
pair_features <- read_tsv("pair_feature_table_AEP.tsv", col_types = cols()) %>%
  mutate(
    dup_type     = factor(dup_type, levels = c("TD","PD","SD","DD")),
    either_opsin = factor(either_opsin,
                          levels = c(FALSE, TRUE),
                          labels = c("non-opsin","opsin"))
  )

# ── Opsin gene IDs ────────────────────────────────────────────────────────────
opsin_ids <- readLines("AEP.txt") %>%
  str_replace("^AEP_", "")

cat("Pairs loaded:    ", nrow(dup_pairs), "\n")
cat("Unique genes:    ", n_distinct(gene_features$gene_name), "\n")
cat("Opsin genes:     ", sum(gene_features$is_opsin), "\n")
cat("Opsin-inv pairs: ", sum(pair_features$either_opsin == "opsin", na.rm=TRUE), "\n")
```

---

## Exploratory Analysis

### Duplication type distribution

The AEP genome contains `r nrow(dup_pairs)` annotated duplicate gene pairs across 
four duplication modes classified by MCScanX. Dispersed duplications (DD) dominate, 
comprising `r scales::percent(mean(dup_pairs$dup_type == "DD"))` of all pairs.

```{r type-counts}
count_summary <- dup_pairs %>%
  count(dup_type) %>%
  mutate(pct = n / sum(n) * 100)

count_summary %>%
  kbl(col.names = c("Type", "n", "%"), digits = 1,
      caption = "Duplicate pair counts by type") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r type-bar, fig.cap="Distribution of duplication types in the Hydra AEP genome."}
ggplot(count_summary, aes(x = dup_type, y = n, fill = dup_type)) +
  geom_col(color = "#0d1117", linewidth = 0.8) +
  geom_text(aes(label = comma(n)), vjust = -0.5,
            size = 3.2, color = "#c9d1d9", fontface = "bold") +
  scale_fill_manual(values = COLORS, guide = "none") +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Duplicate Pair Counts by Type",
       subtitle = "AEP genome | n = 50,364 pairs",
       x = NULL, y = "Number of pairs") +
  theme_hydra
```

### Same-chromosome fraction

Tandem (TD) and proximal (PD) duplicates are expected to reside on the same 
chromosome by definition. Segmental (SD) and dispersed (DD) duplicates may span 
chromosomes.

```{r same-chrom, fig.cap="Fraction of pairs residing on the same chromosome, by duplication type."}
pair_features %>%
  group_by(dup_type) %>%
  summarise(pct_same = mean(same_chrom, na.rm = TRUE) * 100,
            n = n(), .groups = "drop") %>%
  ggplot(aes(x = dup_type, y = pct_same, fill = dup_type)) +
  geom_col(color = "#0d1117", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.0f%%", pct_same)),
            vjust = -0.4, size = 4, color = "#c9d1d9", fontface = "bold") +
  scale_fill_manual(values = COLORS, guide = "none") +
  scale_y_continuous(limits = c(0, 115),
                     labels = function(x) paste0(x, "%")) +
  labs(title = "Same-Chromosome Fraction by Duplication Type",
       x = NULL, y = "% pairs on same chromosome") +
  theme_hydra
```

### Distance between duplicate pairs

Gene ordinal difference (proxy for genomic distance) cleanly separates the 
classification categories — this is partly definitional (TD = adjacent genes), 
but the distribution of PD and DD distances is informative about the scale of 
genomic rearrangement.

```{r distance-ridges, fig.height=5, fig.cap="Distribution of log-transformed ordinal distance between duplicate gene pairs on the same chromosome."}
pair_features %>%
  filter(same_chrom, !is.na(ordinal_diff)) %>%
  mutate(log_od = log10(ordinal_diff + 1)) %>%
  ggplot(aes(x = log_od, y = dup_type,
             fill = dup_type, color = dup_type)) +
  geom_density_ridges(alpha = 0.7, scale = 1.3,
                      quantile_lines = TRUE, quantiles = 2,
                      rel_min_height = 0.01) +
  scale_fill_manual(values = COLORS, guide = "none") +
  scale_color_manual(values = COLORS, guide = "none") +
  labs(title    = "Distance Between Duplicate Pairs (same-chromosome)",
       subtitle = "Vertical line = median; log₁₀(ordinal diff + 1)",
       x = "log₁₀(ordinal diff + 1)", y = NULL) +
  theme_hydra
```

### Gene-level feature distributions

Before modeling, we inspect the raw distributions of each predictor across 
duplication types.

```{r feature-distributions, fig.height=9, fig.cap="Distributions of four genomic predictors across duplication types."}
long_feats <- pair_features %>%
  select(dup_type,
         `Gene length (bp)`       = mean_gene_length,
         `Dist. to nearest gene`  = mean_dist_to_gene,
         `Local gene density`     = mean_local_density,
         `Scaffold length (bp)`   = scaffold_chr_length_dup1) %>%
  pivot_longer(-dup_type, names_to = "feature", values_to = "value") %>%
  filter(!is.na(value))

ggplot(long_feats,
       aes(x = log1p(value), y = dup_type,
           fill = dup_type, color = dup_type)) +
  geom_density_ridges(alpha = 0.65, scale = 1.2,
                      quantile_lines = TRUE, quantiles = 2,
                      rel_min_height = 0.01) +
  facet_wrap(~feature, scales = "free_x", ncol = 2) +
  scale_fill_manual(values = COLORS, guide = "none") +
  scale_color_manual(values = COLORS, guide = "none") +
  labs(title    = "Predictor Distributions by Duplication Type",
       subtitle = "log1p-transformed; vertical line = median",
       x = "log1p(value)", y = NULL) +
  theme_hydra
```

### Opsin gene representation

Of 50 opsin genes, `r sum(pair_features$either_opsin == "opsin", na.rm=TRUE)` 
duplicate pairs involve at least one opsin. This is a small subset, motivating 
the use of Bayesian inference for the opsin-specific analysis.

```{r opsin-counts, fig.cap="Duplication type breakdown for opsin-involving vs non-opsin pairs."}
pair_features %>%
  filter(!is.na(either_opsin)) %>%
  count(either_opsin, dup_type) %>%
  group_by(either_opsin) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = dup_type, y = pct, fill = dup_type)) +
  geom_col(color = "#0d1117", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", pct)),
            vjust = -0.4, size = 3, color = "#c9d1d9") +
  facet_wrap(~either_opsin,
             labeller = labeller(either_opsin = c(
               "opsin"     = "Opsin-involving pairs (n=172)",
               "non-opsin" = "Non-opsin pairs (n=50,192)"
             ))) +
  scale_fill_manual(values = COLORS, guide = "none") +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title    = "Duplication Type Composition: Opsin vs Non-Opsin Pairs",
       subtitle = "Note the elevated TD proportion in opsin-involving pairs",
       x = NULL, y = "% of pairs") +
  theme_hydra
```

### Pairwise correlations among predictors

Checking for multicollinearity before modeling.

```{r correlations, fig.cap="Pairwise Pearson correlations among continuous predictors."}
cor_df <- pair_features %>%
  select(
    gene_length     = mean_gene_length,
    dist_to_gene    = mean_dist_to_gene,
    local_density   = mean_local_density,
    scaffold_length = scaffold_chr_length_dup1
  ) %>%
  drop_na()

cor_mat <- cor(cor_df, method = "pearson")

cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "r") %>%
  ggplot(aes(x = var1, y = var2, fill = r)) +
  geom_tile(color = "#0d1117") +
  geom_text(aes(label = round(r, 2)), color = "#e6edf3", size = 3.5) +
  scale_fill_gradient2(low = "#4ecdc4", mid = "#1c2333",
                       high = "#ff6b6b", midpoint = 0,
                       limits = c(-1, 1), name = "Pearson r") +
  coord_equal() +
  labs(title = "Predictor Correlation Matrix",
       x = NULL, y = NULL) +
  theme_hydra +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### Scaffold-level patterns

Do TD duplications cluster on particular scaffolds?

```{r scaffold-scatter, fig.cap="Relationship between scaffold size and the fraction of TD pairs per scaffold."}
pair_features %>%
  group_by(chrom1 = str_replace(dup1, "\\.G.*", "") %>%
             str_replace("AEP_","")) %>%
  summarise(
    n_pairs    = n(),
    td_pct     = mean(dup_type == "TD") * 100,
    gene_count = first(gc1),
    chrom_size = first(scaffold_chr_length_dup1),
    .groups = "drop"
  ) %>%
  filter(n_pairs >= 10, !is.na(chrom_size)) %>%
  mutate(chrom_label = str_replace(chrom1, "HVAEP", "")) %>%
  ggplot(aes(x = chrom_size / 1e6, y = td_pct,
             size = gene_count, color = gene_count)) +
  geom_point(alpha = 0.85) +
  geom_text(aes(label = chrom_label),
            nudge_y = 0.3, size = 2.8, color = "#8b949e") +
  scale_color_viridis_c(option = "plasma", name = "Gene count") +
  scale_size_continuous(range = c(3, 10), guide = "none") +
  labs(title    = "Scaffold Size vs TD Fraction",
       subtitle = "Bubble size = gene count on scaffold",
       x = "Scaffold length (Mb)", y = "% TD pairs") +
  theme_hydra
```

---

## Data Wrangling Notes

All data processing was performed in R. The key steps and decisions are 
documented here for reproducibility:

1. **Gene ID parsing**: IDs follow the pattern `AEP_HVAEPx.Gnnnnnn`. The 
   chromosome (`HVAEPx`) and gene ordinal (`nnnnnn`) were extracted via regex.

2. **Scaffold lengths**: The GFF3 from NHGRI lacked `##sequence-region` header 
   lines, so scaffold lengths were derived from `max(end)` across all features 
   per scaffold — a conservative but reliable estimate.

3. **Local gene density**: Rather than a scaffold-wide average, we computed 
   a ±500 kb window-based density using `GenomicRanges::countOverlaps()`. This 
   captures regional variation within scaffolds.

4. **Distance to nearest gene**: Computed with `GenomicRanges::distanceToNearest()` 
   on gene-level GRanges, excluding self-hits.

5. **Opsin flag**: 50 opsin genes from lab annotation were matched to GFF3 
   gene names by stripping the `AEP_` prefix. One opsin from the original 51-gene 
   protein dataset was absent from the GFF3 (likely a scaffold not captured 
   in the gene models).

6. **Model dataset**: For the TD vs DD logistic regression, only same-chromosome 
   pairs were retained for TD, and numeric predictors were z-score standardized 
   to allow direct coefficient comparison.

---

## Session Info

```{r session}
sessionInfo()
```
