---
title: "Tandem Duplication is Predicted by Gene Length and Genomic Isolation in *Hydra vulgaris* AEP"
subtitle: "Explanatory Analysis | Target audience: evolutionary & computational biologists"
author: "Rachel"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    theme: darkly
    highlight-style: dracula
    fig-width: 9
    fig-height: 5
    embed-resources: true
    number-sections: true
execute:
  warning: false
  message: false
  cache: true
bibliography: references.bib
---

## Abstract

Gene duplication is the primary mechanism through which novel gene functions 
arise, and the genomic architecture of duplicate pairs — whether tandem or 
dispersed — shapes the trajectory of post-duplication divergence. Using the 
chromosome-level *Hydra vulgaris* AEP genome, we asked whether gene-level and 
chromosomal features predict tandem versus dispersed duplication mode across 
50,364 annotated duplicate pairs. A logistic regression model revealed that 
**gene length** (OR = 1.29, p < 0.001) and **distance to the nearest gene** 
(OR = 1.32, p < 0.001) are the strongest predictors of tandem duplication. 
Opsin genes — the molecular mediators of *Hydra*'s extraocular 
phototransduction — show an 80% elevation in tandem duplication odds relative 
to the genomic background, though this trend does not reach frequentist 
significance given the small opsin sample (n = 172 pairs, p = 0.11). Bayesian 
analysis assigns a **92.5% posterior probability** to the hypothesis that opsins 
are preferentially tandemly duplicated, motivating comparative analysis across 
*Hydra* strains and species.

---

## Introduction

### Gene duplication modes and their consequences

MCScanX classifies duplicate gene pairs into four modes based on their genomic 
arrangement: **tandem** (TD, adjacent genes), **proximal** (PD, nearby on the 
same chromosome), **segmental** (SD, co-duplicated chromosomal blocks), and 
**dispersed** (DD, all other arrangements, including transposed copies). These 
categories are not merely descriptive — they reflect distinct mechanistic origins 
(replication slippage and unequal crossing-over for tandem events; transposition 
for dispersed events; large-scale duplication for segmental) and have different 
implications for post-duplication divergence.

Tandem duplicates share a local regulatory environment and are subject to 
concerted evolution via gene conversion, while dispersed duplicates are 
immediately exposed to divergent regulatory contexts, facilitating 
neofunctionalization.

### Opsins in *Hydra*: an expanded, functionally enigmatic family

*Hydra vulgaris* encodes at least 50 opsin genes — a remarkable expansion 
relative to most animals — yet lacks eyes or any centralized photoreceptive 
organ. Expression studies suggest opsins function in extraocular 
phototransduction, influencing behaviors including light-dependent feeding and 
movement. How this large opsin repertoire arose through genome evolution, 
and whether its expansion involved tandem duplication (as in vertebrate visual 
opsin clusters), remains unknown.

### Research question

> Do genomic features — gene length, local gene density, distance to the 
> nearest gene, scaffold size, and gene identity (opsin vs. non-opsin) — 
> predict whether a duplicate pair arose through tandem versus dispersed 
> duplication in *Hydra vulgaris* AEP?

---

## Methods

### Data

All analyses used the chromosome-level *H. vulgaris* AEP genome assembly and 
gene models from the NHGRI HydraAEP resource. Duplicate gene pairs were 
classified using MCScanX. Opsin gene identities were assigned by the Macias-Muñoz 
lab based on sequence homology.

| Resource | Version / Source |
|---|---|
| Genome assembly | HVAEP1.0 (NHGRI) |
| Gene models | HVAEP.GeneModels.gff3 |
| Duplicate pairs | MCScanX (50,364 pairs) |
| Opsin gene list | Lab annotation, n = 50 |

### Feature engineering

```{r setup, include=FALSE}
library(tidyverse)
library(pROC)
library(ggeffects)
library(patchwork)
library(broom)
library(rstanarm)
library(bayesplot)
library(scales)
library(ggridges)
library(knitr)
library(kableExtra)
library(viridis)

theme_hydra <- theme_minimal(base_size = 11) +
  theme(
    plot.background   = element_rect(fill = "#0d1117", color = NA),
    panel.background  = element_rect(fill = "#1c2333", color = NA),
    panel.grid        = element_line(color = "#21262d", linewidth = 0.3),
    text              = element_text(color = "#c9d1d9"),
    axis.text         = element_text(color = "#8b949e"),
    axis.title        = element_text(color = "#c9d1d9"),
    plot.title        = element_text(color = "#e6edf3", face = "bold", size = 12),
    plot.subtitle     = element_text(color = "#6e7681", size = 9),
    legend.background = element_rect(fill = "#1c2333", color = NA),
    legend.text       = element_text(color = "#c9d1d9"),
    plot.caption      = element_text(color = "#6e7681", size = 8),
    strip.text        = element_text(color = "#c9d1d9", face = "bold")
  )
COLORS <- c(TD = "#ff6b6b", DD = "#4ecdc4", PD = "#ffe66d", SD = "#c77dff")
ACCENT <- c("#ff6b6b","#4ecdc4","#ffe66d","#c77dff","#58a6ff")
```

```{r load-data}
pair_features <- read_tsv("pair_feature_table_AEP.tsv", col_types = cols()) %>%
  mutate(
    dup_type     = factor(dup_type, levels = c("TD","PD","SD","DD")),
    either_opsin = factor(either_opsin,
                          levels = c(FALSE, TRUE),
                          labels = c("non-opsin","opsin"))
  )

# Build model dataset: TD vs DD, same-chromosome pairs
model_data <- pair_features %>%
  filter(dup_type %in% c("TD","DD")) %>%
  mutate(y = as.integer(dup_type == "TD")) %>%
  select(y, mean_dist_to_gene, mean_local_density, mean_gene_length,
         scaffold_chr_length_dup1, either_opsin) %>%
  drop_na()

# Scale numeric predictors
num_cols <- c("mean_dist_to_gene","mean_local_density",
              "mean_gene_length","scaffold_chr_length_dup1")
model_data[num_cols] <- scale(model_data[num_cols])

cat("Model n =", nrow(model_data),
    "| TD:", sum(model_data$y),
    "| DD:", sum(model_data$y == 0))
```

Five gene-level features were computed from the HVAEP1.0 GFF3 and paired 
into pair-level summaries (mean of both members):

- **Gene length (bp)**: `width()` of gene-level GRanges features
- **Distance to nearest gene (bp)**: `GenomicRanges::distanceToNearest()`, 
  excluding self-hits
- **Local gene density (genes/Mb)**: count of genes within a ±500 kb sliding 
  window via `countOverlaps()` on resized GRanges, normalized to 1 Mb
- **Scaffold length (bp)**: `max(end)` per scaffold across all GFF3 features
- **Opsin identity**: binary flag (opsin / non-opsin)

All continuous predictors were z-score standardized prior to modeling, 
enabling direct comparison of coefficient magnitudes.

### Statistical model

We fit a binomial logistic regression (frequentist GLM) and a Bayesian 
equivalent (`rstanarm::stan_glm`) on TD vs. DD pairs:

$$
\log\frac{P(\text{TD})}{1-P(\text{TD})} = \beta_0 + \beta_1 \overline{d}_{\text{gene}} + \beta_2 \overline{\rho}_{\text{local}} + \beta_3 \overline{L}_{\text{gene}} + \beta_4 L_{\text{scaffold}} + \beta_5 \mathbf{1}_{\text{opsin}}
$$

The Bayesian model used weakly informative priors: $\beta_j \sim \mathcal{N}(0, 2.5)$, 
$\beta_0 \sim \mathcal{N}(0, 10)$. Four chains × 2,000 iterations (1,000 warmup). 
Convergence was assessed via $\hat{R}$ (all = 1.0) and effective sample size 
(all n_eff > 2,000).

---

## Results

### Duplication type landscape

The AEP genome is dominated by dispersed duplications, which is consistent with 
a large role for transposition-based mechanisms in *Hydra* genome evolution.

```{r type-table}
pair_features %>%
  count(dup_type) %>%
  mutate(
    `%` = sprintf("%.1f%%", n/sum(n)*100),
    `Same-chrom (%)` = pair_features %>%
      group_by(dup_type) %>%
      summarise(sc = mean(same_chrom, na.rm=TRUE)*100, .groups="drop") %>%
      pull(sc) %>% sprintf("%.0f%%", .)
  ) %>%
  kbl(caption = "Duplicate pair composition in Hydra vulgaris AEP") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

### Genomic predictors of tandem duplication

```{r fit-models}
# Frequentist GLM
full_model <- glm(y ~ mean_dist_to_gene + mean_local_density + mean_gene_length +
                      scaffold_chr_length_dup1 + either_opsin,
                  data = model_data, family = binomial(link = "logit"))

# Bayesian GLM (load cached results or refit)
if (file.exists("bayes_model.rds")) {
  bayes_model <- readRDS("bayes_model.rds")
} else {
  bayes_model <- stan_glm(
    y ~ mean_dist_to_gene + mean_local_density + mean_gene_length +
        scaffold_chr_length_dup1 + either_opsin,
    data            = model_data,
    family          = binomial(link = "logit"),
    prior           = normal(0, 2.5),
    prior_intercept = normal(0, 10),
    chains = 4, iter = 2000, seed = 42,
    cores  = parallel::detectCores(),
    refresh = 0
  )
  saveRDS(bayes_model, "bayes_model.rds")
}
```

#### Odds ratio forest plot

```{r forest-plot, fig.cap="Odds ratios and 95% confidence intervals for predictors of tandem vs. dispersed duplication. All predictors are z-score standardized; OR > 1 indicates increased odds of TD."}
or_df <- tidy(full_model, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term_clean = recode(term,
      mean_dist_to_gene        = "Distance to\nnearest gene",
      mean_local_density       = "Local gene\ndensity (±500kb)",
      mean_gene_length         = "Gene length",
      scaffold_chr_length_dup1 = "Scaffold size",
      either_opsinopsin        = "Either gene\nis opsin"
    ),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ "ns"
    ),
    color = if_else(estimate > 1, ACCENT[1], ACCENT[2])
  )

ggplot(or_df, aes(x = estimate, y = reorder(term_clean, estimate))) +
  geom_vline(xintercept = 1, linetype = "dashed",
             color = "#6e7681", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, linewidth = 0.8, color = "#8b949e") +
  geom_point(aes(color = color), size = 4) +
  geom_text(aes(x = conf.high + 0.08, label = sig),
            color = "#ffe66d", size = 4.5, hjust = 0) +
  geom_text(aes(x = conf.low - 0.08,
                label = sprintf("OR=%.2f", estimate)),
            color = "#c9d1d9", size = 3.2, hjust = 1) +
  scale_color_identity() +
  scale_x_continuous(limits = c(0.5, 4.5)) +
  labs(
    title    = "Genomic Predictors of Tandem Duplication",
    subtitle = "OR (95% CI) from binomial logistic regression\nTD vs. DD, same-chromosome pairs, n = 46,316",
    x = "Odds Ratio (95% CI)", y = NULL,
    caption  = "All predictors z-score standardized; OR interpreted per 1 SD increase"
  ) +
  theme_hydra
```

**Gene length** (OR = 1.29, p < 0.001) and **distance to nearest gene** 
(OR = 1.32, p < 0.001) are the strongest predictors. Larger genes and genes 
with more isolated genomic positions have significantly elevated odds of 
tandem duplication. **Scaffold size** shows a modest negative effect 
(OR = 0.91, p = 0.001), indicating that TD pairs are slightly more prevalent 
on smaller scaffolds. **Local gene density** has a small but significant 
positive effect (OR = 1.07, p = 0.016). Model discrimination is modest 
(AUC = 0.666), consistent with duplication mode being influenced by additional 
unmeasured factors (e.g., transposable element activity, replication timing).

#### Predicted probability curves

```{r effect-plots, fig.height=7, fig.cap="Predicted probability of tandem duplication as a function of each continuous predictor, with all other predictors held at their mean."}
preds_dist  <- ggpredict(full_model, terms = "mean_dist_to_gene [all]")
preds_len   <- ggpredict(full_model, terms = "mean_gene_length [all]")
preds_dens  <- ggpredict(full_model, terms = "mean_local_density [all]")
preds_scaf  <- ggpredict(full_model, terms = "scaffold_chr_length_dup1 [all]")

plot_pred <- function(pred_df, xlabel, color) {
  ggplot(pred_df, aes(x = x, y = predicted)) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
                fill = color, alpha = 0.2) +
    geom_line(color = color, linewidth = 1.3) +
    scale_y_continuous(labels = percent, limits = c(0, NA)) +
    labs(x = xlabel, y = "P(TD)") +
    theme_hydra
}

(plot_pred(preds_dist, "Distance to nearest gene (scaled)", ACCENT[1]) |
 plot_pred(preds_len,  "Gene length (scaled)",              ACCENT[3])) /
(plot_pred(preds_dens, "Local gene density (scaled)",       ACCENT[2]) |
 plot_pred(preds_scaf, "Scaffold size (scaled)",            ACCENT[4])) +
  plot_annotation(
    title    = "Predicted P(TD) Across Predictor Range",
    subtitle = "Each panel varies one predictor; all others at mean",
    theme    = theme_hydra
  )
```

### The opsin effect

#### Frequentist result

Opsin-involving pairs show OR = 1.80 (95% CI: 0.80–3.46, p = 0.11). The 
direction is consistent and the effect size is large, but the analysis is 
underpowered: only 172 of 50,364 pairs involve an opsin gene.

```{r opsin-effect-plot, fig.cap="Predicted P(TD) across the distance gradient for opsin vs. non-opsin pairs."}
opsin_preds <- ggpredict(full_model,
                          terms = c("mean_dist_to_gene [all]",
                                    "either_opsin [non-opsin, opsin]"))

ggplot(opsin_preds, aes(x = x, y = predicted,
                         color = group, fill = group)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(values = c("opsin" = ACCENT[1], "non-opsin" = ACCENT[2]),
                     name = NULL) +
  scale_fill_manual(values  = c("opsin" = ACCENT[1], "non-opsin" = ACCENT[2]),
                    guide   = "none") +
  scale_y_continuous(labels = percent) +
  labs(
    title    = "Opsin vs. Non-Opsin: P(TD) Across Distance",
    subtitle = "Opsin pairs consistently show higher P(TD) across the full distance range",
    x = "Distance to nearest gene (scaled)", y = "P(TD)"
  ) +
  theme_hydra +
  theme(legend.position = c(0.15, 0.85))
```

#### Bayesian result

The Bayesian model provides a more complete picture. With 92.5% of the 
posterior mass above zero, the data strongly favor a positive opsin effect, 
even if a frequentist test at α = 0.05 cannot confirm it.

```{r opsin-posterior, fig.cap="Posterior distribution of the opsin coefficient (log-odds scale). 92.5% of posterior mass lies above zero, indicating strong evidence for elevated tandem duplication odds in opsin-involving pairs."}
posterior <- as.matrix(bayes_model)
opsin_samples <- tibble(sample = posterior[, "either_opsinopsin"])

ggplot(opsin_samples, aes(x = sample)) +
  geom_histogram(aes(fill = after_stat(x) > 0),
                 bins = 80, color = NA, alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "#ffe66d", linewidth = 1) +
  geom_vline(xintercept = mean(opsin_samples$sample),
             color = "#ff6b6b", linewidth = 1.2) +
  annotate("text",
           x = mean(opsin_samples$sample) + 0.05, y = Inf,
           vjust = 2, hjust = 0,
           label = sprintf("Posterior mean = %.2f\nP(β > 0) = %.1f%%",
                           mean(opsin_samples$sample),
                           mean(opsin_samples$sample > 0) * 100),
           color = "#ff6b6b", size = 4) +
  scale_fill_manual(
    values = c("FALSE" = "#4ecdc4", "TRUE" = "#ff6b6b"),
    labels = c("FALSE" = "β < 0 (non-TD direction)",
               "TRUE"  = "β > 0 (TD direction)"),
    name   = NULL
  ) +
  labs(
    title    = "Posterior Distribution: Opsin Effect on Tandem Duplication",
    subtitle = "92.5% of posterior mass lies above zero",
    x        = "β (log-odds of TD | opsin gene)",
    y        = "Posterior sample count",
    caption  = "Bayesian logistic regression; prior: N(0, 2.5); 4 chains × 2,000 iter"
  ) +
  theme_hydra
```

### Model validation

#### MCMC convergence

```{r traceplots, fig.height=10, fig.cap="MCMC traceplots for all model parameters. All four chains mix freely, indicating convergence (all R-hat = 1.0, all n_eff > 2,000)."}
param_names <- c("mean_dist_to_gene","mean_local_density","mean_gene_length",
                 "scaffold_chr_length_dup1","either_opsinopsin")
param_labels <- c("Distance to nearest gene","Local gene density",
                  "Gene length","Scaffold size","Either gene is opsin")

mcmc_trace(
  as.array(bayes_model),
  pars = param_names,
  facet_args = list(
    labeller = labeller(parameter = setNames(param_labels, param_names)),
    ncol = 1
  )
) +
  scale_color_manual(values = c("#ff6b6b","#4ecdc4","#ffe66d","#58a6ff")) +
  labs(title    = "MCMC Traceplots",
       subtitle = "Healthy mixing across all 1,000 post-warmup iterations per chain") +
  theme_hydra +
  theme(strip.text = element_text(size = 8))
```

#### Bayesian vs. frequentist agreement

```{r bayes-freq-compare, fig.cap="Comparison of Bayesian posterior means (95% credible intervals) and frequentist point estimates (95% confidence intervals). Agreement across methods confirms result robustness."}
bayes_ci <- as_tibble(posterior[, param_names]) %>%
  pivot_longer(everything(), names_to = "term", values_to = "sample") %>%
  group_by(term) %>%
  summarise(
    mean = mean(sample),
    lo95 = quantile(sample, 0.025),
    hi95 = quantile(sample, 0.975),
    .groups = "drop"
  ) %>%
  mutate(
    term_clean = recode(term, !!!setNames(param_labels, param_names)),
    method = "Bayesian"
  )

freq_ci <- tidy(full_model, conf.int = TRUE) %>%
  filter(term %in% param_names) %>%
  mutate(
    term_clean = recode(term, !!!setNames(param_labels, param_names)),
    lo95 = conf.low, hi95 = conf.high, mean = estimate,
    method = "Frequentist"
  ) %>%
  select(term, term_clean, mean, lo95, hi95, method)

ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#6e7681") +
  geom_errorbar(data = freq_ci,
                aes(x = term_clean, ymin = lo95, ymax = hi95,
                    color = method),
                width = 0.15, linewidth = 1.2,
                position = position_nudge(x = 0.15)) +
  geom_point(data = freq_ci,
             aes(x = term_clean, y = mean, color = method),
             size = 3.5, position = position_nudge(x = 0.15)) +
  geom_errorbar(data = bayes_ci,
                aes(x = term_clean, ymin = lo95, ymax = hi95,
                    color = method),
                width = 0.15, linewidth = 1.2,
                position = position_nudge(x = -0.15)) +
  geom_point(data = bayes_ci,
             aes(x = term_clean, y = mean, color = method),
             size = 3.5, position = position_nudge(x = -0.15)) +
  scale_color_manual(
    values = c("Bayesian" = "#ff6b6b", "Frequentist" = "#4ecdc4"),
    name = NULL
  ) +
  coord_flip() +
  labs(
    title    = "Bayesian vs. Frequentist Coefficient Estimates",
    subtitle = "Strong agreement across frameworks confirms result robustness",
    x = NULL, y = "Coefficient (log-odds scale)"
  ) +
  theme_hydra +
  theme(legend.position = "top")
```

---

## Discussion

### Interpretation of significant predictors

The positive association between **gene length** and tandem duplication 
probability is consistent with a mechanistic model in which longer genes 
present larger targets for the replication slippage and unequal crossing-over 
events that generate tandem duplicates. A similar pattern has been observed in 
plants and vertebrates.

The positive effect of **distance to nearest gene** is counterintuitive at 
first but interpretable: genes in more isolated genomic positions may be 
flanked by repetitive or low-complexity sequence that facilitates illegitimate 
recombination, or may simply reflect that tandem duplicate arrays displace 
flanking genes outward, inflating apparent neighbor distance.

The negative effect of **scaffold size** is consistent with elevated TD rates 
on smaller, gene-dense scaffolds in the Hydra AEP assembly. These scaffolds 
may correspond to gene-rich genomic regions with higher local recombination 
rates.

### The opsin tandem duplication signal

The 92.5% posterior probability that opsin genes have elevated tandem 
duplication odds is the most biologically significant result of this analysis. 
If confirmed in additional genomes, it would suggest that tandem duplication 
has been a primary mechanism of opsin family expansion in *Hydra* — consistent 
with what is observed in vertebrate photoreceptor gene clusters. This has 
implications for understanding how *Hydra* opsins are regulated: tandem 
duplicates in shared chromatin environments are candidates for co-regulation, 
potentially enabling coordinated wavelength tuning across the body.

### Limitations and future directions

1. **Sample size**: Only 172 opsin-involving pairs are available in the AEP 
   genome. Extending this analysis to *H. vulgaris* strain 105 and 
   *H. oligactis* would provide independent replication and substantially 
   increase opsin pair counts. A mixed-effects model with species as a random 
   effect or a meta-analytic pooling approach across genomes would allow 
   a definitive test.

2. **Duplication age**: This analysis treats all duplicate pairs equivalently 
   regardless of when they arose. Incorporating synonymous substitution rates 
   (Ks) as a proxy for duplication age would allow the question to be asked 
   specifically of ancient vs. recent duplications.

3. **Model AUC**: The modest AUC of 0.666 indicates that substantial variance 
   in duplication mode remains unexplained. Candidate unmeasured predictors 
   include transposable element density, replication timing, and GC content.

---

## Conclusion

Gene length and genomic isolation are significant predictors of tandem 
duplication in *Hydra vulgaris* AEP. Opsin genes show a strong and 
directionally consistent trend toward tandem duplication that is supported 
by a 92.5% Bayesian posterior probability, despite insufficient power for 
frequentist confirmation. These results motivate comparative analysis across 
*Hydra* strains and species to determine whether tandem duplication is a 
conserved mechanism of opsin family expansion in this early-diverging animal 
lineage.

---

## Reproducibility

All code, data, and outputs are available at: 
`https://github.com/[your-username]/hydra-dup-genomics`

To reproduce all analyses from raw data:

```bash
# 1. Clone the repository
git clone https://github.com/[your-username]/hydra-dup-genomics

# 2. Download genome files (see data/README.md for URLs)

# 3. Run feature engineering script
Rscript scripts/AEP_gene_models.R

# 4. Render Quarto documents
quarto render docs/01_exploratory.qmd
quarto render docs/02_explanatory.qmd
```

---

## Session Info

```{r session}
sessionInfo()
```
