---
title: "gene_features"
output: html_document
date: "2026-02-25"
---

## SCRIPT 1: Hydra vulgaris AEP Gene Feature Table

### Install / load packages

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

library(data.table)
library(rtracklayer)   # import() for GFF3
library(GenomicRanges) # GRanges, distanceToNearest, countOverlaps
library(IRanges)       # IRanges (dependency of GenomicRanges)
library(dplyr)
library(readr)
library(stringr)
library(tibble)
```

### Download files Go to "<https://research.nhgri.nih.gov/HydraAEP/download/>" and get the gene model coordinates and unzip it then put in working directory

```{r}
GFF3_FILE <- "HVAEP.GeneModels.gff3.gz"
```

### Parse GFF3 – keep gene-level features only

```{r}
gff_all <- import(GFF3_FILE)
print(table(gff_all$type))
```

### Keep only gene-level features.

Expected: 28,917 records for the AEP assembly.

```{r}
## 
genes <- gff_all[gff_all$type == "gene"]
length(genes)
```

### Pulling scaffold lengths from max(end) per scaffold

Expect 15 chromosomes in this assembly of AEP

```{r}
scaf_max <- as.data.frame(gff_all) %>%
  group_by(seqnames) %>%
  summarise(scaffold_length = max(end), .groups = "drop")
chr_lengths <- setNames(scaf_max$scaffold_length,
                        as.character(scaf_max$seqnames))
length(chr_lengths)
```

### Gene names

AEP gene IDs follow the pattern: HVAEP\<scaffold#\>.G\<6-digit-number\> Example: HVAEP1.G000026

```{r}
#gene_name_vec <- get_attr(genes, "Name")
gene_name_vec <- ifelse(is.na(gene_name_vec),
                        get_attr(genes, "ID"),
                        gene_name_vec)
scaff_names <- as.character(seqnames(genes))
print(head(gene_name_vec))
```

### Base table

```{r}
gene_df <- tibble(
  gene_name           = gene_name_vec,
  scaffold_chr        = scaff_names,
  start               = start(genes),
  end                 = end(genes),
  strand              = as.character(strand(genes)),
  gene_length         = width(genes),
  scaffold_chr_length = chr_lengths[scaff_names]
)
nrow(gene_df)
ncol(gene_df)
```

### Duplicate flag from dup_pairsAEP.tsv

A gene is defined as a "duplicate" if it appears in at least one row of dup_pairsAEP.tsv (That you will need to get from the github and oh I hope its small enough to be stored there).

```{r}
# Something fun I had to learn painfully: dup_pairsAEP.tsv gene IDs carry an "AEP_" prefix (AEP_HVAEP1.G000026), but gene_df gene names do NOT (HVAEP1.G000026). The prefix MUST be stripped before joining, otherwise ALL genes would be flagged as non-duplicates... Yes this took longer than I admit to figure out
dup_pairs <- read_tsv("dup_pairsAEP.tsv", col_types = cols()) 
dup_pairs <- dplyr::rename(dup_pairs, dup_type = type)
#don't ask me why i had to specifically call dplyr every time i used rename but without it, the code wouldn't run :)
dup_pairs <- dup_pairs %>%
    filter(!is.na(dup_type)) %>%
    mutate(
      dup1 = str_replace(dup1, "^AEP_", ""),
      dup2 = str_replace(dup2, "^AEP_", "")
    )
print(count(dup_pairs, dup_type))
```

### Counting which are is_duplicate flag

```{r}
dup_type_lookup <- bind_rows(
  dup_pairs %>% dplyr::select(gene_name = dup1, dup_type),
  dup_pairs %>% dplyr::select(gene_name = dup2, dup_type)
) %>%
  dplyr::distinct(gene_name, .keep_all = TRUE)
```

### Join the dup_type onto gene_df and create is_duplicate

The tidyr replace NA was key here

```{r}
gene_df <- gene_df %>%
  left_join(dup_type_lookup, by = "gene_name") %>%
  mutate(
    dup_type= tidyr::replace_na(dup_type, "unique"),
    is_duplicate= dup_type != "unique"
  )
# Genes with no match in dup_type_lookup get dup_type = "unique" and is_duplicate = FALSE.
gene_df <- gene_df %>%
  left_join(dup_type_lookup, by = "gene_name") %>%
  mutate(
    dup_type= tidyr::replace_na(dup_type, "unique"),
    is_duplicate = dup_type != "unique"
  )
```

### Taking a look at how many duplicates in the AEP genome

```{r}
print(table(gene_df$dup_type))
```

### Opsin flag

Opsin genes are identified by exact match against the list of 50 gene IDs (AEP.txt) that I created by blasting known opsins against the AEP genome (I used Claude to help me remove the blank and header lines) Expect 50 from the sum

```{r}
opsin_ids <- readLines("AEP.txt") %>%
  trimws() %>%
  ## Remove blank lines and any header line
  .[. != "" & . != "AEP"]
gene_df <- gene_df %>%
  mutate(is_opsin = gene_name %in% opsin_ids)
sum(gene_df$is_opsin)
```

### Distance to nearest gene

The sum is the amount of genes with measurable neighbour distance, mine was 28917

```{r}
gene_gr <- makeGRangesFromDataFrame(
  gene_df,
  seqnames.field     = "scaffold_chr",
  start.field        = "start",
  end.field          = "end",
  strand.field       = "strand",
  keep.extra.columns = FALSE
)

dist_hits <- distanceToNearest(gene_gr, ignore.strand = TRUE)

dist_vec <- rep(NA_real_, nrow(gene_df))
dist_vec[queryHits(dist_hits)] <- mcols(dist_hits)$distance

gene_df$dist_near_gene <- dist_vec
sum(!is.na(gene_df$dist_near_gene))
```

### Local gene density (±500 kb window)

I also got help for this one

```{r}
WINDOW <- 500000L
## Expand each gene's range to a centered window.
## fix = "center" anchors the resize on the gene midpoint.
gene_gr_windows <- resize(
  gene_gr,
  width = 2L * WINDOW + width(gene_gr),
  fix   = "center"
)
## Count overlapping genes in each window (ignoring strand).
## This counts the focal gene itself, so we subtract 1.
overlap_counts <- countOverlaps(gene_gr_windows, gene_gr,
                                ignore.strand = TRUE) - 1L
## Convert to genes per Mb: divide by window size in Mb.
## The window is 2*WINDOW bp = 2*WINDOW / 1e6 Mb.
gene_df$local_gene_density <- overlap_counts / (2L * WINDOW / 1e6)
print(summary(gene_df$local_gene_density))
```

### Final table

```{r}
gene_table <- gene_df %>%
  dplyr::select(
    gene_name,
    scaffold_chr,
    start,
    end,
    strand,
    gene_length,
    scaffold_chr_length,
    is_duplicate,
    dup_type,
    is_opsin,
    dist_near_gene,
    local_gene_density
  )
print(head(gene_table, 10))
```

### Save output for Script 2

```{r}
write_csv(gene_table, "hydra_aep_gene_table.csv")
saveRDS(gene_table,   "hydra_aep_gene_table.rds")
```
